### 2022-03-09 unif/unif.tex
more

\blfootnote{In this work I designed the main algorithm, solid oracle and
fingerprint indexing. I also implemented and evaluated the algorithms. Visa
Nummelin proved the completeness of the main algorithm with Alexander Bentkamp's help.
The completeness and termination of solid oracle was proven by me with Alexander
Bentkamp's help.}

@@

Unification is concerned with finding a substitution that makes two terms
equal, for some notion of syntactic equality. Since the invention of Robinson's first-order unification
algorithm \cite{ar-65-resolution}, it has become an indispensable
tool in theorem proving, logic
programming, natural language processing, programming language compilation, and other areas of computer science.

Many of these applications are based on higher-order formalisms and require
higher-order unification. Due to its undecidability and explosiveness,
the higher-order unification problem is considered one of the main
obstacles on the road to efficient higher-order tools.

@@

Jensen and Pietrzykowski's (JP) procedure \cite{jp-76-unif} is the best known
procedure for this purpose (Section \ref{sec:unif:background}). Given two terms to
unify, it first identifies a position where the terms disagree.
Then, in parallel branches of the search tree, it applies suitable substitutions, involving a
variable either at the position of disagreement or above, and repeats this process on
the resulting terms until they are equal or trivially nonunifiable.

Building on the JP procedure, we designed a new procedure (Section
\ref{sec:unif:the-unification-procedure}) with the same completeness guarantees (Section \ref{sec:unif:proof-of-completeness}).
The new procedure addresses many of the issues that are
detrimental to the performance of the JP procedure.
%
First, the JP procedure does not terminate in many cases of obvious
nonunifiability, e.g., for $X \unif \cst{f} \, X$, where $X$ is a nonfunctional
variable and $\cst{f}$ is a function constant. This example also shows that
the JP procedure does not generalize Robinson's first-order procedure gracefully. To address
this issue, our procedure detects whether a unification problem belongs to a
fragment for which unification is decidable and finite complete sets of unifiers (CSUs)
exist.
% Another approach for treating explosiveness of higher-order unification is to
% restrict our attention to some fragment $L$ of $\lambda$-terms that has
% decidable, unary unification problem. We say that the unification problem is
% unary if any two terms belonging to $L$ are either not unifiable or have a most
% general unifier (MGU). 
We call algorithms that enumerate elements of the CSU for such fragments
\emph{oracles}. Noteworthy fragments with oracles are first-order terms,
patterns \cite{tn-93-patterns}, functions-as-constructors
\cite{tl-16-facunif}, and a new fragment 
we present in Section~\ref{sec:unif:solid-oracle}.
%
The unification procedures of Isabelle and Leo-III check whether the unification
problem belongs to a decidable fragment, but we take this idea a step further by
checking this more efficiently and for every subproblem arising during
unification.

@@

% In Section \ref{sec:unif:indexing} we discuss a higher-order extension of fingerprint indexing,
% a technique to filter out terms that are not unifiable with a given
% query term from a set of terms. 
To more efficiently find terms (in a large term set) that are unifiable with a given query term,
we developed a higher-order extension of fingerprint indexing \cite{ss-12-fp-indexing}
(Section \ref{sec:unif:indexing}).
We implemented our procedure, several oracles,
and the fingerprint index in
the Zipperposition prover (Section \ref{sec:unif:implementation}). Since a
straightforward implementation of the JP procedure already existed in
Zipperposition, we used it as a baseline to evaluate the performance of our
procedure (Section \ref{sec:unif:evaluation}). The results show substantial
performance improvements.

% This invited article is an extended version of our FSCD-2020 paper \cite{vbn-21-unif}.
% Most notable extension is the Section \ref{sec:unif:proof-of-completeness}, which gives
% the detailed proof of completeness of our new procedure. In addition, 
% we give proofs for all the unproved statements from the paper, expand the examples 
% and provide more detailed explanations.

@@

\emph{Parameters} and \emph{body} for any term $\param[\seq x]s$ are defined to
be $\seq x$ and $s$ respectively, where $s$ is not a $\lambda$-abstraction. The
\emph{size} of a term is inductively defined as $\text{size}(F) = 1$;
$\text{size}(x) = 1$; $\text{size}(\cst{f}) = 1$; $\text{size}(s\,t) =
\text{size}(s) + \text{size}(t)$; $\text{size}(\lambda x.\, s) = \text{size}(s)
+ 1$. A term is in \emph{head normal form} ({\em hnf}) if it is of the form
$\lambda \overline{x}.\,a\,\overline{t}$, where $a$ is a free variable, bound
variable, or a constant. In this case, $a$ is called the \emph{head} of the
term. Note that this relaxes the condition that the term needs to be in
$\beta$-normal form to determine its head. A term is called flex or rigid if its
head is flex or rigid, respectively. By $\nf{s}{\textsf{h}}$ we denote the term
obtained from a term $s$ by repeated $\beta$-reduction of the leftmost outermost
redex until it is in hnf. Unless stated otherwise, we view terms syntactically,
as opposed to $\alpha\beta\eta$-equivalence classes. The common context
$\mathcal{C}(s,t)$ of two $\eta$-long $\beta$-reduced terms $s$~and~$t$ of the
same type is defined inductively as follows, assuming that $a \not= b$:
$\mathcal{C}(\lambda x.\, s, \lambda y.\, t) = \lambda x.\,
\mathcal{C}(s,\{y\mapsto x\}t)$; $\mathcal{C}(a\,\overline{s}_m,
b\,\overline{t}_n) = \square$; $\mathcal{C}(a\,\overline{s}_m,
a\,\overline{t}_m) = a\,\mathcal{C}(s_1,t_1)\,\ldots\,\mathcal{C}(s_m,t_m)$.
Unless otherwise stated, we take unification constraint $s \unif t$ to be an
unordered pair of two terms of the same type. To ease notation, we do not write
parentheses around application of substitutions to terms (or other objects containing terms);
in other words we shorten $\sigma(\theta(\varrho(s)))$ to $\sigma\theta\varrho \, s$.

\ourpara{Remark}
We use the definition of a CSU from Sect.~\ref{sec:pre:unif} because 
JP's definition of a CSU, which we have adopted in our earlier work \cite{vbn-21-unif}, is flawed.
JP's definition does not employ the notion of auxiliary variables,
but instead requires $\varrho X = \theta\sigma X$ for all variables mapped by $\varrho$.
This is problematic because nothing prevents $\varrho$ from mapping the auxiliary variables.
For example, $\sigma = \{F \mapsto \lambda x y.\> G\>y\}$ is supposed
to be an MGU for $F\>\cst{a}\>\cst{c} \unif F\>\cst{b}\>\cst{c}$.
But for the unifier $\varrho = \{F \mapsto \lambda x y.\> y,\> G \mapsto \lambda x.\> \cst{d}\}$,
without the notion of auxiliary variables,
there exists no appropriate substitution $\theta$
because $\varrho G = \theta\sigma G$ requires $\theta G = \lambda x.\> \cst{d}$
and $\varrho F = \theta\sigma F$ requires $\theta G = \lambda x.\> x$. By declaring $G$ as an auxiliary variable,
we focus only the important variables from the intitial problem (in this case only $F$),
which allows $\sigma$ to be an MGU.

@@

Our approach is to apply substitutions and $\alpha\beta\eta$-normalize terms
lazily. In this context, laziness means that the transitions
\unifrulename{Normalize$_{\alpha\eta}$}, \unifrulename{Normalize$_{\beta}$}, and
\unifrulename{Dereference} partially normalize and partially apply the constructed
substitution just enough to ensure that the heads are the ones we would get if
the substitution was fully applied and the term was fully normalized. 
Additionally, the transitions that modify the constructed substitution,
\unifrulename{OracleSucc} and \unifrulename{Bind}, do not apply that substitution to the
unification pairs directly, but only extend it with a new binding. To support
lazy dereferencing, these rules must maintain the
invariant that all substitutions are idempotent. The invariant is easily
preserved if the substitution $\varrho$ from the definition of
\unifrulename{OracleSucc} and \unifrulename{Bind} is itself idempotent and no variable
mapped by $\sigma$ occurs in $\varrho F$, for any variable $F$ mapped by
$\varrho$.

@@

    where the fresh free variable $G$ as well as all $x_{j_k}$ are of appropriate type.

@@

    appropriate type. Due to indeterminacy of $\overline{y}$, this step is

@@

\begin{exa}
The JP procedure does not terminate on the problem $G \unif \cst{f} \, G$:
\[(G \unif \cst{f} \, G, \makeop{id}) 
\overset{\textsf{Imit}}{\jpunifarrow}
(\cst{f}\,G' \unif \cst{f}^2 \, G', \sigma_1) 
\overset{\textsf{Imit}}{\jpunifarrow}
(\cst{f}^2 \, G'' \unif \cst{f}^3 \, G'', \sigma_2) 
\overset{\textsf{Imit}}{\jpunifarrow}\cdots\]
where $\sigma_1 = \{G \mapsto  \lambda x. \, \cst{f} \, G' \}$
and $\sigma_2 = \{G' \mapsto  \lambda x. \, \cst{f} \, G'' \}\sigma_1$.
By including any oracle that supports the first-order occurs check, such as the
pattern oracle or the fixpoint oracle described in Section \ref{sec:unif:implementation},
our procedure gracefully generalizes first-order unification: 
\[(\{G \unif \cst{f} \, G\}, \makeop{id}) 
\overset{\textsf{OracleFail}}{\newunifarrow}
\bot
\]
\end{exa}

@@

The pragmatic variant of our procedure removes all iteration bindings to enforce finite branching. 
In addition, it
imposes limits 
on the number of bindings applied,
counting the applications of bindings locally, per
constraint, as follows. 
It is useful to distinguish the Huet-style projection cases where
$\alpha_i$ is a base type (called \emph{simple projection}), which always
reduces the problem size, and the cases where $\alpha_i$ is a functional type (called \emph{functional
projection}). We limit the number applications of the following bindings:
functional projections, eliminations, imitations, and identifications.
In addition,
a limit on the total number of applied bindings can be set. 
An elimination binding that removes $k$ arguments counts as $k$
elimination steps.
Thanks to these limits, the pragmatic variant 
terminates.

To fail as soon as any of the limits
is reached, the pragmatic variant employs an additional oracle.
If this oracle determines that the limits are reached and the constraint is of the form
$\lambda \overline{x}.\, F \, \overline{s}_m \unif \lambda \overline{x}.\, G \, \overline{t}_n$, it returns a \emph{trivial unifier} --
a substitution $\{ F \mapsto \lambda \overline{x}_m.\, H, G \mapsto \lambda \overline{x}_n.\, H \}$, where $H$ is a fresh variable; if the limits are reached and
the constraint is flex-rigid, the oracle fails; if the limits are not reached, 
it reports that terms are outside its fragment. The trivial unifier prevents the procedure from failing on
easily unifiable flex-flex pairs.

Careful tuning of each limit optimizes the procedure for a specific class of problems. 
For problems originating from proof assistants, low unification depth usually suffices. However, hard
hand-crafted problems often need
deeper unification.

@@

\begin{theorem}
\label{thm:completeness}
% \looseness=-1
The procedure described in Section \ref{sec:unif:the-unification-procedure}
parameterized by $\mathcal{P}_\mathsf{c}$ is complete, meaning that the
substitutions on the leaves of the constructed tree form a CSU.
More precisely, let $E$ be a multiset of constraints
and let $V$ be the supply of fresh variables provided to the procedure. 
Then for any unifier $\varrho$ of $E$ there exists
a derivation $\left(E,\makeop{id}\right)\longrightarrow^{*}\sigma$ and a
substitution $\theta$ such that for all free variables $X\not\in V$, we have $\varrho X = \theta\sigma X$.
\end{theorem}

@@

In the remaining of this section, we view terms as
$\alpha\beta\eta$-equivalence classes, with the $\eta$-long $\beta$-normal
form as their canonical representative. Moreover, we consider all
substitutions to be fully applied. These assumptions are justified because
all bindings depend only on the head of terms and hence replacing the lazy
transitions $\unifrulename{Normalize}_{\alpha\eta}$,
$\unifrulename{Normalize}_{\beta}$, and $\unifrulename{Dereference}$ by eager
counterparts only affects the efficiency but not the overall behavior of our
procedure.

@@

\begin{exa} To determine the opponent pairs of $\lamx{\cst{g} \, (\cst{g} x)}$
  and $\lamx{\cst{g} \, x}$ we first $\alpha$-rename the second term to 
  $\lam{y}{\cst{g}\,{y}}$. Then, we identify the common context of $\cst{g} \, (\cst{g} x)$ and $\cst{g} \, y$
  as $\cst{g} \, \square $, which leads us to the opponent pair $(\cst{g} \, x, y)$.
\end{exa}

@@

  \begin{exa}
    % \newcommand{\boldboundvar}[1]{\ensuremath{\mathit{\mathbf{#1}}}}
    Following Example 1 in JP's work \cite{jp-76-unif}, we denote
    $\omega$-simple occurrences of parameters using boldface. For example,
    $\lambda xy. \, \cst{f} \, \mathbfit{x} \, \mathbfit{y}$ and $\lamx{\cst{g}
    \, (\lam{y}{\mathbfit{x}\, y})}$ have several $\omega$-simple occurrences.
    In the following two examples, boldface occurrences are \emph{not}
    $\omega$-simple: $\lamx{\cst{g} \, (\lam{y}{\cst{f} \, y \, \mathbfit{x}})},\;
    \lam{xy}{\cst{h} \, (x \, \mathbfit{y})}$. In the first case $x$ is not applied to parameter $y$ and in the second $y$
    occurs as an argument to $x$.
  \end{exa}
  This definition is slightly too restrictive for our purposes.
  It is unfortunate that condition~1 requires $x$ to be applied to
  \emph{all} instead of just some of the bound variables. The JP
  proof would probably work with such a relaxation, and the definition
  would then cover all cases where eager decomposition is possible. 
  However, to reuse the JP lemmas, we stick to the original
  notion of $\omega$-simplicity and introduce the following relaxation:
  \begin{defi}
  An occurrence of a parameter $x$ of term $t$ in the body of $t$
  is \emph{base-simple} if it is $\omega$-simple or both
  \begin{enumerate}
  \item $x$ is of base type, and
  \item this occurrence of $x$ is not in an argument of any parameter
  of $t$.
  \end{enumerate}
  \end{defi}

@@

    $\varrho X\left(\varrho\seq s\right)$ at certain position that depends

@@

    but with the particular exemption: if a prefix variable $u$ 

@@

    We write $<$ for the lexicographic comparison of these tuples.

@@

    Let $\delta$ be
    the Huet-style projection binding:
    \[ \delta = \{a \mapsto \lambda \overline{x}_n. \, x_i \, (F_1 \, \overline{x}_n) \, \ldots \, (F_m \, \overline{x}_n)\} \]
    for fresh variables $F_1,\dots,F_m$.
    This binding is applicable because $b$ is rigid.
    Let $\varrho'$ be the same as $\varrho$ except that we set $\varrho' a = a$
    and for each $1\leq j\leq m$ we set
    \[
      \varrho' F_j = \lambda \overline{x}_n. \, t_j
    \]
    It remains to show that $\ord\varrho'<\ord\varrho$.
    The free weight of $\varrho a$ is the same as 
    the sum of the free weights of $\varrho' F_j$ for $1\leq j\leq m$.
    Thus, the free weight is the same for $\varrho$ and $\varrho'$.
    The bound weight of $\varrho a$ however is exactly $1$ larger than
    the sum of the bound weights of $\varrho' F_j$ for $1\leq j\leq m$
    because of the additional occurrence of $x_k$ in $\varrho a$.
    The exemption for $\omega$-simple occurrences in the definition
    of the bound weight cannot be triggered by this occurrence of $x_k$
    because $m > 0$ and thus $x_k$ is not $\omega$-simple.
    It follows that $\ord\varrho'<\ord\varrho$.
    \end{proof}
    %
    We are now ready to prove the completeness theorem (Theorem \ref{thm:completeness}).
    \begin{proof}
    Let $E$ be a multiset of constraints
    and let $V$ be the supply of fresh variables provided to our procedure. 
    Let $\varrho$ be a unifier of $E$.
    We must show that there exists
    a derivation $(E,\makeop{id})\longrightarrow^{*}\sigma$ and a
    substitution $\theta$ such that for all free variables $X\not\in V$, we have $\varrho X = \theta\sigma X$.

@@

    \begin{description}
    \item[$u$ and $v$ have the same head symbol $a=b$] \hfill
    \begin{enumerate}[beginpenalty=1000]
    \item Suppose that $\varrho_{j}a$ has a parameter with non-base-simple
    occurrence. By one of the induction invariants, $a$ is not an elimination
    variable. 
    Among all non-base-simple occurrences of parameters in $\varrho_{j}a$, 
    choose the leftmost one, which we call $x$.
    This occurrence of $x$ cannot be below another parameter, 
    because having $x$ occur in one of its arguments would make
    that other parameter non-base-simple, contradicting the occurrence of $x$ being leftmost.
    Thus $x$ is neither base-simple nor below another parameter; so $x$ is of functional type.
    Moreover, non-base-simplicity implies non-$\omega$-simplicity.
    Hence, we can apply Lemma~\ref{lem:iteration} (iteration).

@@

      \item Otherwise suppose that $\varrho_j a$, $\varrho_j b$ are in simple
      comparison form (Definition \ref{unif:def:simple-comparison-form}).
      By one of the the induction invariants,
      the free variables occurring in $\varrho_{j}E_{j}$ do not occur in $E_{j}$.
      Thus $\varrho_{j} a \not= a$ and $\varrho_{j} b \not= b$.
      Then Lemma \ref{lem:identification} (identification) applies.

@@

Prehofer's thesis states that PT terminates on $\{s \unif t\}$
if $s$ is linear, $s$ shares no free variables with~$t$, $s$ is strictly solid,
and $t$ is second-order. Together with completeness of PT for preunification,
this result implies that the PT procedure can be used to enumerate finitely many
elements of complete set of preunifiers for this class of terms.

>>>>

\blfootnote{In this work I designed the main algorithm, solid oracle, and
fingerprint indexing. I also implemented and evaluated the algorithms. Visa
Nummelin proved the completeness of the main algorithm with Alexander Bentkamp's help.
The completeness and termination of solid oracle was proven by me with Alexander
Bentkamp's help.}

@@

% Unification is concerned with finding a substitution that makes two terms
% equal, for some notion of syntactic equality. Since the invention of Robinson's first-order unification
% algorithm \cite{ar-65-resolution}, it has become an indispensable
% tool in theorem proving, logic
% programming, natural language processing, programming language compilation, and other areas of computer science.

% Many of these applications are based on higher-order formalisms and require
% higher-order unification. Due to its undecidability and explosiveness,
% the higher-order unification problem is considered one of the main
% obstacles on the road to efficient higher-order tools.

As mentioned in Chapter \ref{ch:intro}, many of the reasoning systems deployed
in various areas of computer science and matemathics are based on variants of
higher-order logic. More often than not, they also require unification of
higher-order terms. Due to its undecidability and explosiveness, the
higher-order unification problem is considered one of the main obstacles on the
road to efficient higher-order tools.

@@

Jensen and Pietrzykowski's (JP) procedure \cite{jp-76-unif} is the best known
procedure for this purpose (Sect.~\ref{sec:unif:background}). Given two terms to
unify, it first identifies a position where the terms disagree.
Then, in parallel branches of the search tree, it applies suitable substitutions, involving a
variable either at the position of disagreement or above, and repeats this process on
the resulting terms until they are equal or trivially nonunifiable.

Building on the JP procedure, we designed a new procedure (Section\ref{sec:unif:the-unification-procedure}) with the same completeness guarantees (Sect.~\ref{sec:unif:proof-of-completeness}).
The new procedure addresses many of the issues that are
detrimental to the performance of the JP procedure.
%
First, the JP procedure does not terminate in many cases of obvious
nonunifiability, e.g., for $X \unif \cst{f} \, X$, where $X$ is a nonfunctional
variable and $\cst{f}$ is a function constant. This example also shows that
the JP procedure does not generalize Robinson's first-order procedure gracefully. To address
this issue, our procedure detects whether a unification problem belongs to a
fragment for which unification is decidable and finite complete sets of unifiers (CSUs)
exist.
% Another approach for treating explosiveness of higher-order unification is to
% restrict our attention to some fragment $L$ of $\lambda$-terms that has
% decidable, unary unification problem. We say that the unification problem is
% unary if any two terms belonging to $L$ are either not unifiable or have a most
% general unifier (MGU). 
We call algorithms that enumerate elements of the CSU for such fragments
\emph{oracles}. Noteworthy fragments with oracles are first-order terms,
patterns \cite{tn-93-patterns}, functions-as-constructors
\cite{tl-16-facunif}, and a new fragment 
we present in Sect.~\ref{sec:unif:solid-oracle}.
%
The unification procedures of Isabelle and Leo-III check whether the unification
problem belongs to a decidable fragment, but we take this idea a step further by
checking this more efficiently and for every subproblem arising during
unification.

@@

% In Sect.~\ref{sec:unif:indexing} we discuss a higher-order extension of fingerprint indexing,
% a technique to filter out terms that are not unifiable with a given
% query term from a set of terms. 
To more efficiently find terms (in a large term set) that are unifiable with a given query term,
we developed a higher-order extension of fingerprint indexing \cite{ss-12-fp-indexing}
(Sect.~\ref{sec:unif:indexing}).
We implemented our procedure, several oracles,
and the fingerprint index in
the Zipperposition prover (Sect.~\ref{sec:unif:implementation}). Since a
straightforward implementation of the JP procedure already existed in
Zipperposition, we used it as a baseline to evaluate the performance of our
procedure (Sect.~\ref{sec:unif:evaluation}). The results show substantial
performance improvements.

% This invited article is an extended version of our FSCD-2020 paper \cite{vbn-21-unif}.
% Most notable extension is the Sect.~\ref{sec:unif:proof-of-completeness}, which gives
% the detailed proof of completeness of our new procedure. In addition, 
% we give proofs for all the unproved statements from the paper, expand the examples 
% and provide more detailed explanations.

@@

\emph{Parameters} and \emph{body} for any term $\param[\seq x]s$ are defined to
be $\seq x$ and $s$ respectively, where $s$ is not a $\lambda$-abstraction. The
\emph{size} of a term is inductively defined as $\text{size}(F) = 1$;
$\text{size}(x) = 1$; $\text{size}(\cst{f}) = 1$; $\text{size}(s\,t) =
\text{size}(s) + \text{size}(t)$; $\text{size}(\lambda x.\, s) = \text{size}(s)
+ 1$. A term is in \emph{head normal form} ({\em hnf}) if it is of the form
$\lambda \overline{x}.\,a\,\overline{t}$, where $a$ is a free variable, a bound
variable, or a constant. In this case, $a$ is called the \emph{head} of the
term. Note that this relaxes the condition that the term needs to be in
$\beta$-normal form to determine its head. A term is called flex or rigid if its
head is flex or rigid, respectively. By $\nf{s}{\textsf{h}}$ we denote the term
obtained from a term $s$ by repeated $\beta$-reduction of the leftmost outermost
redex until it is in hnf. Unless stated otherwise, we view terms syntactically,
as opposed to $\alpha\beta\eta$-equivalence classes. The common context
$\mathcal{C}(s,t)$ of two $\eta$-long $\beta$-reduced terms $s$~and~$t$ of the
same type is defined inductively as follows, assuming that $a \not= b$:
$\mathcal{C}(\lambda x.\, s, \lambda y.\, t) = \lambda x.\,
\mathcal{C}(s,\{y\mapsto x\}t)$; $\mathcal{C}(a\,\overline{s}_m,
b\,\overline{t}_n) = \square$; $\mathcal{C}(a\,\overline{s}_m,
a\,\overline{t}_m) = a\,\mathcal{C}(s_1,t_1)\,\ldots\,\mathcal{C}(s_m,t_m)$.
Unless otherwise stated, we take unification constraint $s \unif t$ to be an
unordered pair of two terms of the same type. To ease notation, we do not write
parentheses around application of substitutions to terms (or other objects containing terms);
in other words we shorten $\sigma(\theta(\varrho(s)))$ to $\sigma\theta\varrho \, s$.

\ourpara{Remark}
We use the definition of a CSU from Sect.~\ref{sec:pre:unif} because 
JP's definition of a CSU, which we have adopted in our earlier work \cite{vbn-21-unif}, is flawed.
JP's definition does not employ the notion of auxiliary variables,
but instead requires $\varrho X = \theta\sigma X$ for all variables mapped by $\varrho$.
This is problematic because nothing prevents $\varrho$ from mapping the auxiliary variables.
For example, $\sigma = \{F \mapsto \lambda x y.\> G\>y\}$ is supposed
to be an MGU for $F\>\cst{a}\>\cst{c} \unif F\>\cst{b}\>\cst{c}$.
But for the unifier $\varrho = \{F \mapsto \lambda x y.\> y,\> G \mapsto \lambda x.\> \cst{d}\}$,
without the notion of auxiliary variables,
there exists no appropriate substitution $\theta$
because $\varrho G = \theta\sigma G$ requires $\theta G = \lambda x.\> \cst{d}$
and $\varrho F = \theta\sigma F$ requires $\theta G = \lambda x.\> x$. By declaring $G$ as an auxiliary variable,
we focus only on the important variables from the initial problem (in this case only $F$),
which allows $\sigma$ to be an MGU.

@@

Our approach is to apply substitutions and $\alpha\beta\eta$-normalize terms
lazily. In this context, laziness means that the transitions
\unifrulename{Normalize$_{\alpha\eta}$}, \unifrulename{Normalize$_{\beta}$}, and
\unifrulename{Dereference} partially normalize and partially apply the constructed
substitution just enough to ensure that the heads are the ones we would get if
the substitution was fully applied and the term was fully normalized. 
Additionally, the transitions that modify the constructed substitution,
\unifrulename{OracleSucc} and \unifrulename{Bind}, do not apply that substitution to the
unification pairs directly, but only extend it with a new binding. To support
lazy dereferencing, these rules must maintain the
invariant that all substitutions are idempotent. The invariant is easily
preserved if the substitution $\varrho$ from the definitions of
\unifrulename{OracleSucc} and \unifrulename{Bind} is itself idempotent and no variable
mapped by $\sigma$ occurs in $\varrho F$, for any variable $F$ mapped by
$\varrho$.

@@

    where the fresh free variable $G$ as well as all $x_{j_k}$ are of appropriate types.

@@

    appropriate types. Due to indeterminacy of $\overline{y}$, this step is

@@

\begin{exa}
The JP procedure does not terminate on the problem $G \unif \cst{f} \, G$:
\[(G \unif \cst{f} \, G, \makeop{id}) 
\overset{\textsf{Imit}}{\jpunifarrow}
(\cst{f}\,G' \unif \cst{f}^2 \, G', \sigma_1) 
\overset{\textsf{Imit}}{\jpunifarrow}
(\cst{f}^2 \, G'' \unif \cst{f}^3 \, G'', \sigma_2) 
\overset{\textsf{Imit}}{\jpunifarrow}\cdots\]
where $\sigma_1 = \{G \mapsto  \lambda x. \, \cst{f} \, G' \}$
and $\sigma_2 = \{G' \mapsto  \lambda x. \, \cst{f} \, G'' \}\sigma_1$.
By including any oracle that supports the first-order occurs check, such as the
pattern oracle or the fixpoint oracle described in Sect.~\ref{sec:unif:implementation},
our procedure gracefully generalizes first-order unification: 
\[(\{G \unif \cst{f} \, G\}, \makeop{id}) 
\overset{\textsf{OracleFail}}{\newunifarrow}
\bot
\]
\end{exa}

@@

The pragmatic variant of our procedure removes all iteration bindings to enforce finite branching. 
In addition, it
imposes limits 
on the number of bindings applied,
counting the applications of bindings locally, per
constraint, as follows. 
It is useful to distinguish the Huet-style projection cases where
$\alpha_i$ is a base type (called \emph{simple projection}), which always
reduces the problem size, and the cases where $\alpha_i$ is a functional type (called \emph{functional
projection}). We limit the number of applications of the following bindings:
functional projections, eliminations, imitations, and identifications.
In addition,
a limit on the total number of applied bindings can be set. 
An elimination binding that removes $k$ arguments counts as $k$
elimination steps.
Thanks to these limits, the pragmatic variant 
terminates.

To fail as soon as any of the limits
is reached, the pragmatic variant employs an additional oracle.
If this oracle determines that the limits are reached and the constraint is of the form
$\lambda \overline{x}.\, F \, \overline{s}_m \unif \lambda \overline{x}.\, G \, \overline{t}_n$, it returns a \emph{trivial unifier} --
a substitution $\{ F \mapsto \lambda \overline{x}_m.\, H, G \mapsto \lambda \overline{x}_n.\, H \}$, where $H$ is a fresh variable; if a limit is reached and
the constraint is flex-rigid, the oracle fails; if the limits are not reached, 
it reports that terms are outside its fragment. The trivial unifier prevents the procedure from failing on
easily unifiable flex-flex pairs.

Careful tuning of each limit optimizes the procedure for a specific class of problems. 
For problems originating from proof assistants, a low unification depth usually suffices. However, hard
hand-crafted problems often need
deeper unification.

@@

\begin{theorem}
\label{thm:completeness}
% \looseness=-1
The procedure described in Sect.~\ref{sec:unif:the-unification-procedure}
parameterized by $\mathcal{P}_\mathsf{c}$ is complete, meaning that the
substitutions on the leaves of the constructed tree form a CSU.
More precisely, let $E$ be a multiset of constraints
and let $V$ be the supply of fresh variables provided to the procedure. 
Then for any unifier $\varrho$ of $E$ there exists
a derivation $\left(E,\makeop{id}\right)\longrightarrow^{*}\sigma$ and a
substitution $\theta$ such that for all free variables $X\not\in V$, we have $\varrho X = \theta\sigma X$.
\end{theorem}

@@

In the remainder of this section, we view terms as
$\alpha\beta\eta$-equivalence classes, with the $\eta$-long $\beta$-normal
form as their canonical representative. Moreover, we consider all
substitutions to be fully applied. These assumptions are justified because
all bindings depend only on the heads of terms and hence replacing the lazy
transitions $\unifrulename{Normalize}_{\alpha\eta}$,
$\unifrulename{Normalize}_{\beta}$, and $\unifrulename{Dereference}$ by eager
counterparts only affects the efficiency but not the overall behavior of our
procedure.

@@

\begin{exa} To determine the opponent pairs of $\lamx{\cst{g} \, (\cst{g} x)}$
  and $\lamx{\cst{g} \, x}$, we first $\alpha$-rename the second term to 
  $\lam{y}{\cst{g}\,{y}}$. Then, we identify the common context of $\cst{g} \, (\cst{g} x)$ and $\cst{g} \, y$
  as $\cst{g} \, \square $, which leads us to the opponent pair $(\cst{g} \, x, y)$.
\end{exa}

@@

  \begin{exa}[JP E1]
    % \newcommand{\boldboundvar}[1]{\ensuremath{\mathit{\mathbf{#1}}}}
    Term
    $\lambda xy. \, \cst{f} \, \mathbfit{x} \, \mathbfit{y}$ has two and term
    $\lamx{\cst{g} \, (\lam{y}{\mathbfit{x}\, y})}$ has one $\omega$-simple
    occurrence, denoted in boldface. In the following two examples, boldface
    occurrences are \emph{not} $\omega$-simple: $\lamx{\cst{g} \,
    (\lam{y}{\cst{f} \, y \, \mathbfit{x}})},\; \lam{xy}{\cst{h} \, (x \,
    \mathbfit{y})}$. In the first case $x$ is not applied to parameter $y$ and
    in the second $y$ occurs as an argument to $x$. 
  \end{exa}
  This definition is slightly too restrictive for our purposes.
  It is unfortunate that condition~1 requires $x$ to be applied to
  \emph{all} instead of just some of the bound variables. The JP
  proof would probably work with such a relaxation, and the definition
  would then cover all cases where eager decomposition is possible. 
  However, to reuse the JP lemmas, we stick to the original
  notion of $\omega$-simplicity and introduce the following relaxation:
  \begin{defi}
  An occurrence of a parameter $x$ of term $t$ in the body of $t$
  is \emph{base-simple} if it is $\omega$-simple or both
  \begin{enumerate}
  \item $x$ is of base type, and
  \item this occurrence of $x$ is not in an argument of any parameter
  of $t$.
  \end{enumerate}
  \end{defi}

@@

    $\varrho X\left(\varrho\seq s\right)$ at a certain position that depends

@@

    but with a particular exemption: if a prefix variable $u$ 

@@

    We write $<$ for the lexicographic comparison of these triples.

@@

    Let $\delta$ be
    the Huet-style projection binding:
    \[ \delta = \{a \mapsto \lambda \overline{x}_n. \, x_i \, (F_1 \, \overline{x}_n) \, \ldots \, (F_m \, \overline{x}_n)\} \]
    for fresh variables $F_1,\dots,F_m$.
    This binding is applicable because $b$ is rigid.
    Let $\varrho'$ be the same as $\varrho$ except that we set $\varrho' a = a$
    and for each $1\leq j\leq m$ we set
    \[
      \varrho' F_j = \lambda \overline{x}_n. \, t_j
    \]
    It remains to show that $\ord\varrho'<\ord\varrho$.
    The free weight of $\varrho a$ is the same as 
    the sum of the free weights of $\varrho' F_j$ for $1\leq j\leq m$.
    Thus, the free weight is the same for $\varrho$ and $\varrho'$.
    The bound weight of $\varrho a$ however is exactly $1$ larger than
    the sum of the bound weights of $\varrho' F_j$ for $1\leq j\leq m$
    because of the additional occurrence of $x_k$ in $\varrho a$.
    The exemption for $\omega$-simple occurrences in the definition
    of the bound weight cannot be triggered by this occurrence of $x_k$
    because $m > 0$ and thus $x_k$ is not $\omega$-simple.
    It follows that $\ord\varrho'<\ord\varrho$.
    \end{proof}
    %
    We are now ready to prove the completeness theorem (Theorem \ref{thm:completeness}).
    \begin{proof}
    Let $E$ be a multiset of constraints
    and let $V$ be the supply of fresh variables provided to our procedure. 
    Let $\varrho$ be a unifier of $E$.
    We must show that there exist
    a derivation $(E,\makeop{id})\longrightarrow^{*}\sigma$ and a
    substitution $\theta$ such that for all free variables $X\not\in V$, we have $\varrho X = \theta\sigma X$.

@@

    \begin{description}
    \item[$u$ and $v$ have the same head symbol $a=b$] \hfill
    \begin{enumerate}[beginpenalty=1000]
    \item Suppose that $\varrho_{j}a$ has a parameter with a non-base-simple
    occurrence. By one of the induction invariants, $a$ is not an elimination
    variable. 
    Among all non-base-simple occurrences of parameters in $\varrho_{j}a$, 
    choose the leftmost one, which we call $x$.
    This occurrence of $x$ cannot be below another parameter, 
    because having $x$ occur in one of its arguments would make
    that other parameter non-base-simple, contradicting the occurrence of $x$ being leftmost.
    Thus $x$ is neither base-simple nor below another parameter; so $x$ is of functional type.
    Moreover, non-base-simplicity implies non-$\omega$-simplicity.
    Hence, we can apply Lemma~\ref{lem:iteration} (iteration).

@@

      \item Otherwise suppose that $\varrho_j a$, $\varrho_j b$ are in simple
      comparison form (Definition \ref{unif:def:simple-comparison-form}).
      By one of the induction invariants,
      the free variables occurring in $\varrho_{j}E_{j}$ do not occur in $E_{j}$.
      Thus $\varrho_{j} a \not= a$ and $\varrho_{j} b \not= b$.
      Then Lemma \ref{lem:identification} (identification) applies.

@@

Prehofer's thesis states that PT terminates on $\{s \unif t\}$
if $s$ is linear, $s$ shares no free variables with~$t$, $s$ is strictly solid,
and $t$ is second-order. Together with completeness of PT for preunification,
this result implies that the PT procedure can be used to enumerate finitely many
elements of a complete set of preunifiers for this class of terms.

Prehofer focused on the preunification problem as he remarks that the resulting
flex-flex pairs are ``intricate" \cite[Sect.~5.2.2]{cp-95-unifphd}. We discovered that these flex-flex
pairs actually have an MGU, allowing us to solve the full unification problem,
rather than the preunification problem. Moreover, we lift the order restriction
imposed by Prehofer: We identify a class of terms called \emph{solid} which
requires that all arguments of free variables are either bound variables (of
arbitrary type) or ground (arbitrary-order) terms of base type.
